### 2020-08-07 thesis/thesis.tex
[thesis] Checkpoint in writing thesis

\polishtitle{Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak{} w językach programowania}
\englishtitle{Coalgebraic effects and their cohandlers \fmlinebreak{} in programming languages}
\polishabstract{
    \ldots
}
\englishabstract{
    \ldots
}
\author{Mateusz Urbańczyk}
\advisor{dr Maciej Piróg}
\date{1 września 2020}                     % Data zlozenia pracy
% Dane do oswiadczenia o autorskim wykonaniu
% \transcriptnum {291480}                     % Numer indeksu
% \advisorgen    {dr. Macieja Piróga} % Nazwisko promotora w dopelniaczu
%%%%%

%%%%% WLASNE DODATKOWE PAKIETY
% \usepackage{graphicx,listings,tikz}
\usepackage{syntax}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{mathalfa}
\usepackage{textcomp}
\usepackage{stmaryrd}
\usepackage{tikz-cd}

\usepackage[backend=bibtex]{biblatex}
\addbibresource{mybib.bib}
%
%%%%% WŁASNE DEFINICJE I POLECENIA
%
\theoremstyle{definition} \newtheorem{definition}{Definition}[chapter]
\theoremstyle{remark} \newtheorem{remark}[definition]{Observation}
\theoremstyle{plain} \newtheorem{theorem}[definition]{Theorem}
\theoremstyle{plain} \newtheorem{lemma}[definition]{Lemma}
%\renewcommand \qedsymbol {\ensuremath{\square}}

@@

In this thesis we discuss issues with excessive generality of algebraic effects,
and propose a solution without coalgebraic means along with presentation of
experimental programming language Freak, which is a language with (co) algebraic
effects and (co) handlers, where implementaiton is based on Continutation Passing
Style translation.

\section{Problem Statement}

Algebraic effects, while not being a new concept, in the recent years have
received a lot of attention~\cite{effects-bibliography}, both from theoretical
and practical side.

They give developer necessary tools for declarative programming and writing
programs in a way to strive from exposing implementation details by defining
computational effects as an API\@. In the same time, they preserve a lot of
flexibility and have strong theoretical background.

In fact, that very flexibility can be troublesome. Allowing developer for too
much freedome may lead to undesired behaviour and incorrect programs, for that
very reason, we develop type systems, to detect errors before they occur.

Multi-shot resumptions in effect handlers are a powerful tool to describe
fairly complex logic in a concise way. That being said, they give rise to
issues that would not occur in standard control flow, especially when composing
various effects together.

We will explain this in greater details and think about possible ways to address
the issue of flexibility of effects in the next sections.

TODO actually point out which sections you talk about or maybe even put example
here

@@

We start with Chapter~\ref{chapter:background}, which provides a background
about effects, defines algebraic effects in categorical setting as well as point
out dual coalgebraic effects and cohandlers, finishing with showing related work
in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
to approach issue of excesive generality,
Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
examples along with usage guide, and then
Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
operational semantics and CPS translation. In next one,
Chapter~\ref{chapter:implementation}, implementation details are revealed.
We conclude in Chapter~\ref{chapter:conclusion} by stating what are the possible
augmentations, that are intended to be made in the future.

@@

Since the rapid development of computational theory in 1930s by A. Turing,
K. Godel and A. Church, we have a well-established notion of what can and what
cannot be done through algorithmic means, which we can almost directly translate
to being computable by our machines. Through next years we have developed
mainstream languages that are used almost everywhere, with a great success.

Under these circumstances one may pose a question, why do we still bother with
development of languages theory, since so much has been done already. Is there
anything that drives us towards further research? Indeed, one active branch
revolves around equational theory to asses equality of two programs, which we
know that in general setting is undecidable. Proof methods may include extensional,
contextual or logical equality. However, there is no doubt that these formal
ways of reasoning about programs, while being crucial for assessing correctness,
do not bring direct benefits for everyday use cases, as they are rarely accessible
by a common developer.

Other branch of languages theory, that we shall investigate more in this thesis,
is about taming complexity of programs. Various methods of static analysis has
been developed for various use cases, most notably, type systems.Thanks to strong
and static type systems along with their implementations, we have solid tools to
work efficiently on functions that are pure. That being said, we claim that the
core complexity of programs comes from side effects, or more generally, computational
effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects. One of the ways
to model them, can be done through monads~\cite{monads-wadler, moggi}. However,
they were found to be, to say the least, cumbersome to work with when the number
of different effects increases. It is perhaps not a coincidence that many
functional programming languages do not have them, because of the non-composable
nature of them, or at least not composable in their implementations.

Let's put the following functions

@@

where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
What in case of adding one more effect, $f''$ and $g''$?

It turns out, that it becomes complex and unpleasent to combine two functors
together to form a new monad, and for this purpose, monad
transformers~\cite{monad-transformers}arose in Haskell. Most of the common
languages avoid this by not expressing computational effects in the type system,
and instead one may think about functions as being implicitly embed in a Kleisli
Category over a functor T, where T is a hidden signature over all possible side
effects that occur in our program.

TODO Think about describing what is Kleisli Category

Not only we would like to bring back effects to our type
system~\cite{type-and-effect}, but also do it in a way that is composable. This
is where algebraic effects comes to the rescue.  From theoretical point of view,
we need to develop equational theory about our effects to assert correctness of
our langauge as well as to have the right hammer to reason about our programs.
That is the point where we would like to introduce to unfamiliar reader a notion
of algebraic effects.

@@

Algebraic effects can be thought as an public interface for computational effects.
Declarative approach allow us to write programs in which the actual semantic of
source code is dependent on handler that defines the meaning of a subset of effects.

This is really an incredible feature from practical point of view, as we may substitute
logic depending on execution environment. As an example, fetching for resources
can behave differently as we run tests, debug our code, or run it on production.
In the same manner, they neatly allow us to abstract over implementation details.

@@

TODO maybe example with comparison for these two approaches could be useful here?

That being said, it's only one particular issue that algebraic effects address.
In effects approach, handler may also drop the resumption or invoke it more than
once. There are many other great sources for getting familiarized with
effects~\cite{handlers-tutorial, programming-in-eff, koka-tutorial}, so we will
omit further explainations. More examples can be found in
Chapter~\ref{chapter:co-effectful-programming}.

@@

Algebraic effects can be described via operational means, however,
for the purpose of presenting the duality between algebra and coalgebra,
we allow ourselves to wander a bit deeper into category theory and describe
effects from denotational point of view.

@@

    \begin{definition}

@@

    \end{definition}

    \begin{definition}
    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
    signature $ \Sigma $ and morphisms into folds over trees.

    \end{definition}

    \begin{definition}

@@

    \end{definition}

    \begin{definition}

    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given
    by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
    $\mathVar{\Sigma_{T}-equations}$. For clarity, we will usually omit T subscript.

    \end{definition}

    \begin{definition}

@@

    \begin{center}
        $ {\llbracket op_{i} \rrbracket}_I : P_{i} \times{} {|I|}^{A_{i}} \rightarrow |I| $
    \end{center}
    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
    which sends variables into projections from environment and terms into
    map composition over each subterm.

    \end{definition}

    \begin{definition}

@@

    \begin{center}
    % \begin{tikzcd}[row sep=huge, column sep=large]
    %     {|M|}^k \arrow[rr, '\phi', bend right] \arrow[rr, '\rrbracket{} X|l', bend left] &  & {|M|}
    % \end{tikzcd}
    \end{center}

    TODO this diagram doesn't work

    \begin{center}
    % \begin{tikzcd}
    % A \arrow[rd] \arrow[r, '\phi'] & B \\
    % & C
    % \end{tikzcd}
    \end{center}

    \end{definition}

    \begin{definition}

@@

    \end{center}
    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
    Universality means that for every algebra $ \beta : F B \longrightarrow B $
    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
    extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram
    below commutes:

    TODO we can also say it's an initial algebra over initial object.

    \begin{center}
    \begin{tikzcd}[row sep=huge, column sep=large]
    % F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
    % F B \arrow[r, "\beta"']                                       & B                                &
    \end{tikzcd}
    \end{center}

    \end{definition}

    \begin{definition} \textit{Free model} is just a model which is free algebra.

    \end{definition}

    \begin{lemma}{Free models form monads}

    \end{lemma}

    \begin{definition}

    Let L, M be models of a theory T. A $ \mathVar{T-homomorphism}$
    $\phi : L \rightarrow M$ is a map such that the following diagram commutes:

    TODO finish definition

    \end{definition}

@@

    We could end by stating this definition, and everything else would follow
    directly from duality. However, expanding definitions is going to give us
    better intuition, as well as give more solid ground for implementation.

@@

    Derive from Models duality

@@

@@

    Induction and Coinduction

    Induction is a way of constructing new structures. Recursion is a way of
    folding inductively defined structure in an terminating way. Recursive functions
    should shrink the argument in each call, meaning that it eventually ends up
    terminating in a base case.

    Coinduction is literally a dual notion to induction. We \textit{observe} possibly
    infinite structures, by doing deconstruction. Corecursion, however, is a way
    of productively defining new, possibly enlarged structures. Due to infinity,
    the evaluation should be lazy, whereas in induction it may be eager.

    TODO cite Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors

@@

    Reader that focuses on pragmatism may pose a question, why do we even want
    to reason about infinite structures? They never appear in practice! In fact,
    it's very common to operate on never-ending transition systems or streams of
    data without, wher efinitary means of reasoning are of no use, as we can’t
    expect an end to stream!

    Simple example of the duality can be expressed through induction over finite
    list and coinductive observation of infinite streams.

    Doing the latter may involve modification of the internal state of the
    machine that is generating the infinite streams, or in more concrete
    scenario, alternation of the external resource that is providing us the data.

    From categorical standpoint, coinduction is formed over a final coalgebra,
    where corecursion is the mediator between any coalgbra into a final one,
    and coinductive proof principle corresponds to uniqueness of the mediator.
    Recall that coalgebraic effects arises from a particular type of a final
    coalgebra, namely, cofree coalgebra.

    TODO Write a coinductive proof for divergent terms using coinductive proof
    principle. Introdcution to bisim, page 34,35.

\subsection{Coinduction Coalgebra something section}

    TODO Rethink this section, back by examples

    This is one of the cases where interaction with external resource multiple
    times, or more specifically in case of algebraic effects, invoking resumption
    more than once, may lead to unexpected behaviour that would not be expected
    in standard control flow.

    Let's consider the following example

    TODO rethink easter eg
    \begin{verbatim}
    handle
        let fh <- do Open "praise.txt" in
        let c <- do Choice () in
        if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool") ;
        close fh
    with {
        return x -> [x] |
        choose () k -> return (append (k true) (k false)) |
    }
    \end{verbatim}

    Where we open file, and based on nondeterministic choice, we write to file,
    and then close it. This piece of code looks harmless, however, as we invoke
    resumption for the second time, we are attempting to write to a closed file,
    and then close it once again. This captures the excessive generality of
    effects, and is the issue that we would like to address with coalgebraic
    effects.

@@

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
    effects are implemented through generators~\cite{one-shot}, using built-in
    feature of sending value when doing \textit{yield} operation. Resumptions in
    handlers that are sending value, are one-shot and tail-recursive, therefore
    we do not need to handle coalgebraic part, at the cost of flexibility.

    Except from Links language~\cite{handlers-cps}, on which the implementation
    is based, there are currently many other alternatives available. One may take
    a look at Frank~\cite{frank}, which provides a support for multihandlers,
    Koka~\cite{leijen-koka}, Helium~\cite{helium} or Eff~\cite{eff}. Except from
    separate languages, many libraries arose for existing ones like Haskell,
    Idris, Scala or Multicore OCaml.

@@

    Ahman and Bauer~\cite{runners-in-action}

@@

TODO smth smt initialization before finalization.

It can be seen, that issues lies with the fact, that interacting with coalgebras
may lead to change of their internal state system. Examples include change
of state in NFA, taking next element in infinite stream or a closing file.

One of the ways to approach this problem, caused by extensive generality of
effects, is extension of the calculus with Coalgebraic effects, also named
coeffects. Difference is that resumptions in coeffects handlers are one-shot
only, which means that continuation can be invoked at most once.

\begin{lemma}
    Problem of detection whether resumption is called only once is undecidable.
\end{lemma}

\begin{proof}
    Follows directly from Rice's Theorem, as checking whether function was called
    is a nontrivial semantic property of a program.
\end{proof}

@@

One approach is to dynamically during execution check against contract that
continuation may only be called once. In this setting we are sure that our
program will not accidentaly go into wrong state, however by definition we lack
static analysis to prevent from errors before they occur.

@@

Solve the issue through introduction of linear type system.

TODO cite LINEAR USAGE OF STATE

In fact, there is a bijective translation between algebraic effects and linear
type theory, and \textit{every monad embeds in a linear state monad}.

TODO Explain it can get complex and unfamiliar for users.

>>>>

\polishtitle{Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak{} w językach programowania}
\englishtitle{Coalgebraic effects and their cohandlers \fmlinebreak{} in programming languages}
\polishabstract{
    \ldots
}
\englishabstract{
    \ldots
}
\author{Mateusz Urbańczyk}
\advisor{dr Maciej Piróg}
\date{1 września 2020}                     % Data zlozenia pracy
% Dane do oswiadczenia o autorskim wykonaniu
\transcriptnum{291480}                     % Numer indeksu
\advisorgen{dr. Macieja Piróga} % Nazwisko promotora w dopelniaczu
%%%%%

% \usepackage[utf8]{inputenc}
% \usepackage{graphicx,listings,tikz}
\usepackage{syntax}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{mathalfa}
\usepackage{textcomp}
\usepackage{stmaryrd}
\usepackage{bussproofs}
\usepackage{tikz-cd}
\usetikzlibrary{babel}

% Disable certain warnings
% chktex-file 18

\usepackage[backend=bibtex]{biblatex}
\addbibresource{mybib.bib}

\theoremstyle{definition} \newtheorem{definition}{Definition}[chapter]
\theoremstyle{remark} \newtheorem{remark}[definition]{Observation}
\theoremstyle{plain} \newtheorem{theorem}[definition]{Theorem}
\theoremstyle{plain} \newtheorem{lemma}[definition]{Lemma}
%\renewcommand \qedsymbol {\ensuremath{\square}}

@@

In this thesis we discuss issues with excessive generality of algebraic effects,
and propose a solution without coalgebraic means along with presentation of
experimental programming language Freak, which is a language with (co) algebraic
effects and (co) handlers, where implementation is based on Continuation Passing
Style translation.

\section{Problem Statement}\label{sec:problem-statement}

Algebraic effects, while not being a new concept, in the recent years have
received a lot of attention~\cite{effects-bibliography}, both from the theoretical
and practical side.

They give developer necessary tools for declarative programming and writing
programs in a way to strive from exposing implementation details, by defining
computational effects as an API\@ in our code. In the same time, they preserve
a lot of flexibility and have a strong theoretical background.

In fact, that very flexibility can be troublesome. Allowing developer for too
much freedom may lead to undesired behaviour and incorrect programs, for this
reason, we develop type systems, to detect errors before they occur.

Multi-shot resumptions in effect handlers and possibility to drop resumption
are powerful tools to describe fairly complex logic in a concise way. That being
said, they give rise to issues that would not occur in standard control flow,
especially when composing various effects together. Unwanted interaction with
external resources multiple times or not invoking finalisation code are cases
that may occur while using algebraic effects and handlers. Let's consider the
following example:

\begin{verbatim}
handle
    let fh <- do Open "praise.txt" in
    let c <- do Choice () in
    if c then do Write (fh, "Guy Fieri") else do Write (fh, "is cool") ;
    close fh
with {
    return x -> [x] |
    choose () k -> return (append (k true) (k false)) |
}
\end{verbatim}

In the above code we open file, and based on nondeterministic choice, we write
to the file, and then close it. This piece of code looks harmless, however, as
we invoke resumption for the second time, we are attempting to write to the
closed file, and then close it once again. In fact, dropping resumption is also
considered harmful, as we would not attempt to close the file.

This captures the excessive generality of effects, and is the issue that we
would like to address with coalgebraic effects.

@@

In the following chapter we provide a background
about effects, define algebraic effects in categorical setting as well as point
out dual coalgebraic effects and cohandlers, finishing with showing related work
in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
to approach issue of excessive generality,
Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
examples along with usage guide, and then
Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
operational semantics and CPS translation. In next one,
Chapter~\ref{chapter:implementation}, implementation details are revealed.
We conclude in Chapter~\ref{chapter:conclusion} by stating what are the possible
augmentations, that are intended to be made in the future.

@@

Since the rapid development of computational theory in 1930s by A. Turing,
K. Godel and A. Church, we have a well-established notion of what can and what
cannot be done through algorithmic means, which we can almost directly translate
to being computable by our machines. Through next years we have developed
mainstream languages that are used almost everywhere, with great success.

Under these circumstances one may pose a question, why do we still bother with
development of languages theory, since so much has been done already. Is there
anything that drives us towards further research? Indeed, one active branch
revolves around equational theory to assess equality of two programs, which we
know that in the general setting is undecidable. Proof methods may include
extensional, contextual or logical equality. However, there is no doubt that
these formal ways of reasoning about programs, while being crucial for assessing
correctness, do not bring direct benefits for everyday use cases, as they are
rarely accessible by a common developer.

Other branch of languages theory, that we shall investigate more in this thesis,
is about taming complexity of programs. Various methods of static analysis has
been developed for various use cases, most notably, type systems. Thanks to strong
and static type systems along with their implementations, we have solid tools to
work efficiently on functions that are pure. That being said, we claim that the
core complexity of programs comes from side effects, or more generally, computational
effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects. One of the ways
to model them, can be done through monads~\cite{monads-wadler, moggi}. However,
they were found to be, to say the least, cumbersome to work with when the number
of different effects increase. It is perhaps not a coincidence that many
functional programming languages do not have them, because of the non-composable
nature of them, or at least not composable in their implementations.

\noindent
Let's put the following functions

@@

\noindent
where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
What in case of adding one more effect, $f''$ and $g''$?

It turns out, that it becomes complex and unpleasant to combine two functors
together to form a new monad, and for this purpose, monad
transformers~\cite{monad-transformers} arose in Haskell. Most of the common
languages avoid this by not expressing computational effects in the type system,
and instead one may think about functions as being implicitly embed in a Kleisli
Category over a functor T~\cite{kleisli-ncatlab}, where T is a hidden signature
over all possible side effects that occur in our program.

\noindent
TODO Reconsider reference to ncatlab on Kleisli Cat, often it's not the best resource

Not only we would like to bring back effects to our type
system~\cite{type-and-effect}, but also do it in a way that is composable. This
is where algebraic effects comes to the rescue. From theoretical point of view,
we need to develop equational theory about our effects to assert correctness of
our language as well as to have the right hammer to reason about our programs.
From practical side, we need to have the way of taming computational effects
in our programs. That is the point where we would like to introduce to unfamiliar
readers a notion of algebraic effects.

@@

Algebraic effects can be thought of as an public interface for computational effects.
Declarative approach allow us to write programs in which the actual semantic of
source code is dependent on handler that defines the meaning of a subset of effects.

This is really an incredible feature from practical point of view, as we may substitute
logic depending on the execution environment. As an example, fetching for resources
can behave differently as we run tests, debug our code, or run it on production.
In the same manner, they neatly allow us to abstract over implementation details.

@@

\noindent
TODO maybe example with comparison for these two approaches could be useful here?

That being said, it's only one particular issue that algebraic effects address.
Their handlers may also drop the resumption or invoke it more than
once. There are many other great sources for getting familiarized with
effects from practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
so we will omit further explanations. More examples can be found in
Chapter~\ref{chapter:co-effectful-programming}.

@@

Algebraic effects can be described via operational means, however,
for the purpose of presenting the duality between algebra and coalgebra,
we allow ourselves to wander a bit deeper into category theory and describe
effects from denotational point of view~\cite{bauer-what-is-algebraic}.

@@

    % == begin Signature ==
    \begin{definition}

@@

    \end{definition}
    % == end Signature ==

    % == begin Sigma-terms ==
    \begin{definition}
    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
    signature $ \Sigma $ and morphisms into folds over trees.

    \end{definition}
    % == end Sigma-terms ==

    % == begin Sigma-equation ==
    \begin{definition}

@@

    \end{definition}
    % == end Sigma-equation ==

    % == begin Algebraic Theory ==
    \begin{definition}

    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given
    by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
    $\mathVar{\Sigma_{T}-equations}$.

    \end{definition}
    % == end Algebraic Theory ==

    % == begin Interpretation ==
    \begin{definition}

@@

    \begin{center}
        $ {\llbracket op_{i} \rrbracket}_I : P_{i} \times{} {|I|}^{A_{i}} \rightarrow |I| $
    \end{center}

    \noindent
    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
    which sends variables into projections from environment and terms into
    map composition over each subterm.

    \end{definition}
    % == end Interpretation ==

    % == begin Model ==
    \begin{definition}

@@

    \begin{center}
    \begin{tikzcd}[row sep=huge, column sep=large]
        {|M|}^k \arrow[rr, "\llbracket{} X|r \rrbracket{}"', bend right] \arrow[rr, "\llbracket{} X|l \rrbracket{}", bend left] &  & {|M|}
    \end{tikzcd}
    \end{center}

    \end{definition}
    % == end Model ==

    % == begin homomophism ==
    \begin{definition}

    Let L, M be models of a theory T. $ \mathVar{T-homomorphism}$
    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
    $op_{i}$ in T the following diagram commutes:

    \begin{center}
    \begin{tikzcd}
        {|L|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{L}"] \arrow[dd, "\varphi{}^{ar_{i}}"'] &  & {|L|} \arrow[dd, "\varphi{}"] \\
                                              &  &                                     \\
        {|M|}^{ar_{i}} \arrow[rr, "{\llbracket{} op_{i} \rrbracket{}}_{M}"']                 &  & {|M|}
    \end{tikzcd}
    \end{center}

    \noindent
    We denote here $A^{n}$ as $\mathVar{n-ary}$ product of $A$.

    \end{definition}
    % == end homomophism ==

    % == begin Free algebra ==
    \begin{definition}

@@

    \noindent
    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
    Universality means that for every algebra $ \beta : F B \longrightarrow B $
    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
    extending f, i.e.\ a unique morphism of $ \mathcal{C}$ for which the diagram
    below commutes:

    \begin{center}
    \begin{tikzcd}[row sep=huge, column sep=large]
    F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
    F B \arrow[r, "\beta"']                                       & B                                &
    \end{tikzcd}
    \end{center}

    \end{definition}
    % == end Free algebra ==

    \begin{definition} \textit{Free model} is just a model that is free algebra.
    \end{definition}

    \begin{lemma}{Free models form monads}
    \end{lemma}
    \begin{proof}
        Let $M$ be a free model of an algebraic theory $T$ in category $\mathcal{C}$. We
        have an endofunctor $ F_{T}$, which takes objects into free model and
        maps to unique homomorphisms for which the following diagram commutes:

        \begin{center}
        \begin{tikzcd}
        X \arrow[rr, "\eta_{X}"] \arrow[dd, "f"'] &  & F_{T} (X) \arrow[dd, "\overline{f}"] \\
                                                  &  &                                     \\
        Y \arrow[rr, "\eta_{Y}"']                 &  & F_{T} (Y)
        \end{tikzcd}
        \end{center}

        \noindent
        which immediately gives us $\eta$ natural transformation for a monad.
        We can now define $\mu{}$ for a monad as a unique morphism for which
        the diagram commutes:

        \begin{center}
        \begin{tikzcd}[row sep=huge, column sep=large]
            F_{T} (X) \arrow[r, "\eta_{F_{T} (X)}"] \arrow[rd, "id"'] & F_{T} (F_{T} (X)) \arrow[d, "\mu{}"] \\
                                                         & F_{T} (Y)
        \end{tikzcd}
        \end{center}

        \noindent
        $F_{T}$ is an endofunctor, therefore it sends $X$ into an object in $\mathcal{C}$.
        From any object we have a unique map sending an object into model,
        which is $\eta_{F_{T} (X)}$. From free property of our model we get
        a unique map $\mu{}$ such that the diagram commutes, therefore we have
        a monad $(F_{T}, \eta{}, \mu{})$.

    \end{proof}

    % == begin handler ==
    \begin{definition} \textit{Handler} is a $\mathVar{T-homomorphism}$ between
    free models

        \begin{center}
        $ H : {| F_{T} (X) |} \rightarrow {| F_{T'} (X') |} $
        \end{center}

    \noindent
    that is, a map between carriers that preserves the structure of theory $T$.

    \end{definition}
    % == end handler ==

Concluding from the theory that we have built, algebraic effects are just free
models over computation trees, where the signature is the set of effects, and
equational theory is definining rules of rewriting expressions with effects.
The latter is usually irrelevant from implementation perspective, but may be
important for reasoning about programs.

Effect handlers on the other hand, are transformations from one computation
tree over given signature of effects into another one, where handler may serve
effects and also propagate new ones.

@@

    We could end by stating this definition, and everything else would follow
    directly from duality. However, expanding definitions is going to give us
    better intuition, as well as give more solid ground for implementation.
    For this part, we are going to assume we operate in \textbf{Set} category.

@@

    We have defined interpretation of an operation as a map:

    \begin{center}
        $ {\llbracket op \rrbracket}_M : {|I|}^{A} \rightarrow |I|^{P} $
    \end{center}

    \noindent
    Let's now take a look on how it dualizes. In \textbf{Set}, we can think about
    $A^{B}$ exponentials as $\mathVar{B-ary}$ coproducts over A, where for each
    argument we select one result. Every arrow in opposite category is reverted,
    thus coproducts becomes products, our morphism is reverted and model becomes
    comodel (\textit{world}). Putting this together, we obtain the following map:

    \begin{center}
        $ {\llbracket op \rrbracket}^W : {|I|} \times {P} \rightarrow  {|I|} \times {A} $
    \end{center}

    \noindent
    which is called a \textit{cooperation}. Meaning of that morphism, is that
    based on coalgebra carrier and a parameter, we obtain new state of the
    coalgebra and a value generated by coalgebra.

@@

\noindent
TODO \\
- coalg effects \\
- cohandlers

@@

    Induction is a way of constructing new structures. Recursion is a way of
    folding inductively defined structure in a terminating way. Recursive
    functions (not to be confused with recursive computability class), should
    shrink the argument in each call, meaning that it eventually ends up
    terminating in a base case.

    Coinduction is literally a dual notion to induction. We \textit{observe} possibly
    infinite structures, by doing deconstruction. Corecursion, is a way
    of productively defining new, possibly enlarged structures. Due to infinity,
    the evaluation should be lazy, whereas in induction it may be eager.

@@

    From categorical standpoint, coinduction is formed over a final coalgebra,
    where corecursion is the mediator between any coalgebra into a final one,
    and coinductive proof principle corresponds to the uniqueness of the mediator.
    Recall that coalgebraic effects arises from a particular type of a final
    coalgebra, namely, cofree coalgebra.

    \noindent
    TODO Not sure which paper to cite here for a deeper dive into the upper
    topic. Book \textit{Initial Algebras, Terminal Coalgebras, and the Theory of Fixed Points of Functors}
    by Adámek, Milius and S. Moss from which I have studied is no longer available
    on the internet, as it was a draft under construction.

    Reader that focuses on pragmatism may pose a question, why do we even want
    to reason about infinite structures? They never appear in practice! In fact,
    it's very common to operate on never-ending transition systems or streams of
    data, where finitary means of reasoning are of no use, as we can’t expect
    an end to stream.

    Operating on coinductive structures may involve modification of the internal
    state of the machine that is generating the infinite streams, or in more
    concrete scenario, alternation of the external resource that is providing
    us the data.

    For a better understanding and intuition, we shall illustrate now the
    difference based on inductive and coinductive relation. Recall that the set
    $\Lambda$ of $\mathVar{\lambda-terms}$ is given by the following grammar:

    \begin{center}
        $ e ::= x \;|\; \lambda x.e \;|\; e_{1} e_{2} $
    \end{center}

    \noindent
    where $\Lambda^{0} \in \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
    meaning, those terms have no free variables. Let's now define an inductive,
    and coinductive predicate. Relation
    $\Downarrow{} \subseteq \Lambda^{0} \times \Lambda^{0}$ (convergence) for
    call-by-value $\mathVar{\lambda-calculus}$ is defined as follows:

    \begin{center}
        \AxiomC{}
        \UnaryInfC{$ \lambda x.e \Downarrow{} \lambda x.e $}
        \DisplayProof{}
        \quad\quad
        \AxiomC{$ e_{1} \Downarrow{} \lambda x.e_{0}$}
        \AxiomC{$e_{0}[e_{2}/x] \Downarrow{} e'$}
        \BinaryInfC{$ e_{1} e_{2} \Downarrow{} e' $}
        \DisplayProof{}
    \end{center}

    \noindent
    which means that $ \Downarrow{} $ is the \textit{smallest} predicate
    \textit{closed forward} under the above rules. Relation can be inductively
    built from empty set and then in a fixed-point manner we can add more terms.
    Let's now see a relation $ \Uparrow{} \subseteq \Lambda^{0}$ of
    \textit{divergent} terms, which is coinductively defined as follows:

    \begin{center}
        \AxiomC{$ e_{1} \Uparrow{} $}
        \UnaryInfC{$ e_{1} e_{2} \Uparrow{} $}
        \DisplayProof{}
        \quad\quad
        \AxiomC{$ e_{1} \Downarrow{} \lambda x.e_{0}$}
        \AxiomC{$e_{0}[e_{2}/x] \Uparrow{}$}
        \BinaryInfC{$ e_{1} e_{2} \Uparrow{}$}
        \DisplayProof{}
    \end{center}

    \noindent
    which means that $\Uparrow$ is the \textit{greatest} predicate
    \textit{closed backwards} under the above rules, relation can be coinductively
    obtained by starting with $\Lambda_{0}$, and then by removing elements that
    do not fit the rules.

    \noindent
    For more detailed introduction into coinduction and other examples,
    see~\cite{jacobs-rutten, sangiorgi-intro}.

    \noindent
    TODO maybe also write coinductive proof based on two relations above?

@@

    Except from Links language~\cite{handlers-cps}, on which the implementation
    is based, there are currently many other alternatives available. One may take
    a look at Frank~\cite{frank}, which provides support for multihandlers,
    Koka~\cite{leijen-koka}, Helium~\cite{helium} or Eff~\cite{eff}. Except from
    separate languages, many libraries arose for existing ones like Haskell,
    Idris, Scala or Multicore OCaml.

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
    effects are implemented through generators~\cite{one-shot}, using built-in
    feature of sending value when doing \textit{yield} operation. Resumptions in
    handlers that are sending value are one-shot and tail-recursive, therefore
    we do not need to handle the coalgebraic part, at the cost of flexibility.

@@

    Work that is the closest related to topic of the thesis, is work done by
    Danel Ahman and Andrej Bauer~\cite{runners-in-action}, as well as their
    implementation of a $\lambda_{coop}$ language~\cite{coop}, which is a
    calculus that ensures linear usage of resources as well as execution of
    code handling finalisation.

\noindent
TODO Consider those\\
- From comodels to coalgebras: State and arrays.\\
- Tensors of comodels and models for operational semantics.\\
- Stateful runners of effectful computations.

@@

It can be seen, that issues lie with the fact, that interacting with coalgebras
may lead to change of their internal state system. Examples include change of
state in NFA, taking next element in an infinite stream or a closing file.
Finalization must occur after initialization, thus we conclude that resumption
should be called at least once. However, modifying state twice the same way can
also lead to unwanted behaviour, as it was seen in
Section~\ref{sec:problem-statement}, therefore it should be invoked exactly
once.

\begin{lemma}
    Problem of detection whether resumption is called only once is undecidable.
\end{lemma}
\begin{proof}
    Follows directly from Rice's Theorem, as checking whether given function is
    going to be called is a nontrivial semantic property of a program.
\end{proof}

One of the ways to approach this problem, caused by extensive generality of
effects, is the introduction of dual Coalgebraic effects, also named coeffects,
for which we shall discuss various methods of implementation.

@@

One approach is to dynamically check against contract that continuation may only
be called once. In this setting we are sure that our program will not accidentally
go into wrong state, however by definition we lack static analysis to prevent
errors before they occur.

@@

Imposition of a single call of resumption, can be thought in a way that continuation
is `consumable', such that it's not available after one takes use of it. Putting
it in that way, it immediately reminds of linear logic~\cite{linear-logic},
where premises of linear implications are no longer of use in conclusion, and
in case of linear type theory, binding variable from expression A, disallows
use of variables in A. Implementation of coalgebraic effects might be simplified
in languages that already support linear type system, such as Rust~\cite{rust}.
